---
title: "AFL Browne-Low Raw Stats, Analytics, and Pre Processing"
output: html_notebook
---
This project was conducted by Thomas Johnstone, Richard, and Luke Mason, with supervision from RMIT Lecturer Dr Haydar Demirhan Bayesian Statistics 2022.


This R Notebook is responsible for pre processing the AFL stats so that they can be consumed by the JAGS model.
## Imports
```{r}
library(tidyr)
```


## Load in the training set
```{r}
raw_stats = read.csv("../../data/raw_stats.csv")
```


Create the list of round names
```{r}
# Create initial list of valid round names.
valid_round_names <- c()

# Create 23 round names.
for (num in 1:23) {
  # Append round names to valid round name list
  valid_round_names <- c(valid_round_names, paste("R", num, sep=""))
}

# Display.
valid_round_names
```

```{r}

# Set of column names that are not numerical, or do not contribute to the calculation of brownlow points
drops <- c("gameId","gameNumber", "Subs", "year", "team", "X..Played", "playerId", "displayName", "round")

# Drop these column names from the raw stats into filtered dataframe.
filtered_df <- raw_stats[ , !(names(raw_stats) %in% drops)]

# Aggregate all fields in the filtered data frame, but group them by the list of values in field columns:
# - playerId; so that all games of the player are aggregated together by mean
# - displayName; so that we can have the name of the player during data analysis
# - round; so that the player's games are not aggregated into 1 row, but instead 1 row per round.
agg = aggregate(filtered_df, by=list(raw_stats$playerId, raw_stats$displayName, raw_stats$round), FUN = mean)

# Remove rows who's round is not in the list of valid round names
# This is because the finals rounds exist after round 22 and there are no brownlow points assigned to players during these games, which makes these rows redundant.
df1 <- agg[ agg$Group.3 %in% valid_round_names, ]

stat_cols = c("Behinds", "Bounces", "Brownlow.Votes", "Clangers", "Clearances", "Contested.Marks", "Contested.Possessions",
    "Disposals", "Frees", "Frees.Against", "Goal.Assists", "Goals", "Handballs",
    "Hit.Outs", "Inside.50s", "Kicks", "Marks", "Marks.Inside.50", "One.Percenters", "Rebounds",
    "Tackles", "Uncontested.Possessions", "X..Played")

df2 = pivot_longer(df1, names_to="stat", values_to="value", cols=all_of(stat_cols))
df3 = pivot_wider(df2, id_cols = c("Group.1", "Group.2", "stat"), names_from = "Group.3")

# Order the columns
col_order = c("Group.1", "Group.2", "stat", valid_round_names)
df4 <- df3[, col_order]

# Rename columns
names(df4)[names(df4) == 'Group.1'] <- 'playerId'
names(df4)[names(df4) == 'Group.2'] <- 'displayName'

df4
```




```{r}
# Rename columns
names(df1)[names(df1) == 'Group.1'] <- 'playerId'
names(df1)[names(df1) == 'Group.2'] <- 'displayName'
names(df1)[names(df1) == 'Group.3'] <- 'round'

stat_df <- df1[order(df1$playerId, df1$displayName, df1$round),]
stat_df
```

```{r}
filtered_df2 <- filtered_df[ , !(names(filtered_df) %in% c("rounds"))]
# Aggregate all fields in the filtered data frame, but group them by the list of values in field columns:
# - playerId; so that all games of the player are aggregated together by mean
# - displayName; so that we can have the name of the player during data analysis
# - round; so that the player's games are not aggregated into 1 row, but instead 1 row per round.
per_player_agg = aggregate(filtered_df, by=list(raw_stats$playerId, raw_stats$displayName), FUN = mean)

# Rename columns
names(per_player_agg)[names(per_player_agg) == 'Group.1'] <- 'playerId'
names(per_player_agg)[names(per_player_agg) == 'Group.2'] <- 'displayName'

ppa <- per_player_agg[order(per_player_agg$playerId, per_player_agg$displayName),]
ppa
```


```{r}
all_stats <- raw_stats[ raw_stats$round %in% valid_round_names, ]
all_stats
```

# Pre processing the % average performance for each player per game
```{r}
library(dplyr)
library(tidyr)

# Convert the indices to integers
convert_list_stoi <- function(list_str) {
  redundant_rows <- c()
  for(idx in 1:length(list_str)) {
    redundant_rows <- c(redundant_rows, strtoi(list_str[idx], base=10))
  }
  
  return(redundant_rows)
}

# Creates the performance dataset but from x number of games and optionally without many 0 votes.
create_performance_dataset <- function(df, number_of_games, balanced = FALSE){


  # Group the rows into games, by game id
  top <- df %>% group_by(gameId) %>% summarise()
  gameIds <- as.list(head(top, number_of_games))
  game_stats <- df[df$gameId %in% unlist(gameIds),]
  
  # Drop the redundant columns
  drops <- c("team", "round",	"gameNumber")
  game_stats <- game_stats[ , !(names(game_stats) %in% drops)]
  
  stat_cols <- c(
    "Behinds",
    "Bounces",
    "Clangers",
    "Clearances",
    "Contested.Marks",
    "Contested.Possessions",
    "Disposals",
    "Frees", 
    "Frees.Against",
    "Goal.Assists",
    "Goals",
    "Handballs",
    "Hit.Outs",
    "Inside.50s",
    "Kicks",
    "Marks",
    "Marks.Inside.50",
    "One.Percenters",
    "Rebounds",
    "Tackles",
    "Uncontested.Possessions"
    )
  
  xPlayed <- "X..Played"
  label <- "Brownlow.Votes"

  
  # Column names.
  x <- c(stat_cols, c(xPlayed, label, "displayName", "year", "gameId", "playerId"))
  
  # The new data frame to create.
  new_df <- data.frame(matrix(ncol = length(x), nrow = 0))
  
  # Assign column names.
  colnames(new_df) <- x
  
  # For each group
  for (id in unlist(gameIds)) {
    game <- game_stats[game_stats$gameId == id,]
    
      # Removed 40 of 41 the players with 0 brownlow votes.
    if (balanced) {
      # Read in the stats of the players
      game$id <- 1:nrow(game)
     
      # Removed 40 of 41 the players with 0 brownlow votes.
      indices <- unlist(game[game$Brownlow.Votes == 0,]$id)

      # Select all but the first
      indices <- indices[2:length(indices)]

      # Remove the selected rows, to be left with a single 0, 1, 2, and 3 brownlow votes row
      row_indices <- unlist(game[-indices,]$id)
    } else {
        row_indices <- unlist(1:nrow(game))
    }
  
    # For each player in the group, get the rest of the players in that group and aggregate their stats
    for (row_index in row_indices) {
  
      # Get the player row
      player <- game[c(row_index),]
  
      # Get the rest of the rows seperately.
      rest_of_players <- game[-c(row_index),]
  
      # Drop columns because they appeared again
      player_filtered <- player[,names(player) %in% stat_cols]
      rest_of_players_filtered <- rest_of_players[ , names(rest_of_players) %in% stat_cols]
  
      # Get the mean value of each column, then add value and stats columns
      player_agg <- data.frame(colMeans(rest_of_players_filtered))
      player_agg$stat <- rownames(player_agg)
      player_agg$value <- player_agg$colMeans.rest_of_players_filtered.
      
      # Shift the columns to rows but has a column header this time for both dataframes.
      player_filtered <- pivot_longer(player_filtered, names_to="stat", values_to="value", cols=all_of(stat_cols))
      
      # Order by stat name so that each dataset can marry up their values correctly
      player_filtered = player_filtered[order(player_filtered$stat),]
      player_agg = player_agg[order(player_agg$stat),]
      
      # Evaluate the performance of the player vs the rest of the players in the game
      performance <- data.frame(performance = (player_filtered$value / player_agg$value) - 1, stat = player_agg$stat)
  
      # Replace the Inf values with 1 = 100%, 0 would not represent the player having beat all the other players.
      is.na(performance)<-sapply(performance, is.infinite)
      performance[is.na(performance)]<-1
  
      # Shift rows to columns
      performance <- pivot_wider(data=performance, names_from="stat", values_from="performance")
  
      # Add columns to help post processing
      performance[xPlayed] <- as.matrix(player[xPlayed])[1] / 100
      performance[label] <- as.matrix(player[label])[1]
      performance["displayName"] <- as.matrix(player["displayName"])[1]
      performance["year"] <- as.matrix(player["year"])[1]
      performance["gameId"] <- as.matrix(player["gameId"])[1]
      performance["playerId"] <- as.matrix(player["playerId"])[1]
      
      # print(player)
      # Add the row to the new data frame.
      new_df[nrow(new_df) + 1,] <- performance[1,]
      
    }
    print(paste("Finished Game ", id))
  }
  
  return(new_df)
}
```

### Create more balanced performance dataset
```{r}
# Read in the stats of the players
all_stats <- read.csv("../../data/all_stats.csv")

# Make the train set balanced
train_performance = all_stats[all_stats$year != "2021",]
balanced_performance_df = create_performance_dataset(train_performance, 10000, TRUE)

# Make the test set be all of the games of 2021 so all can be tallied up.
test_performance = all_stats[all_stats$year == "2021",]
test_performance_df = create_performance_dataset(test_performance, 10000, FALSE)

# Make the test set be all of the games of 2021 so all can be tallied up.
test_balanced_performance = all_stats[all_stats$year == "2021",]
test_blanaced_performance_df = create_performance_dataset(test_balanced_performance, 10000, TRUE)
```

# Save balanced split performance datasets
```{r}
# Write train and test sets to files.
name = "performance"
write.csv(balanced_performance_df, paste("../../data/train_", name, ".csv", sep=""), row.names=FALSE)
write.csv(test_performance_df, paste("../../data/test_", name, ".csv", sep=""), row.names=FALSE)
write.csv(test_blanaced_performance_df, paste("../../data/test_balanced_", name, ".csv", sep=""), row.names=FALSE)
```

### Create the performance dataset
```{r}
performance_df = create_performance_dataset(all_stats, 100)
performance_df
```

## Save datasets
```{r}
write.csv(stat_df, "../../data/stats_per_unique_round.csv", row.names=FALSE)
write.csv(per_player_agg, "../../data/stats_per_career.csv", row.names=FALSE)
write.csv(all_stats, "../../data/all_stats.csv", row.names=FALSE)
write.csv(performance_df, "../../data/raw_performance.csv", row.names=FALSE)
write.csv(balanced_performance_df, "../../data/performance.csv", row.names=FALSE)
```







